Towards an Ecological Trait-data Standard

### 2019-03-21 manuscript/manuscript.Rmd
edits on manuscript, part I

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1]  \footnotemark[2]
- name: David Fichtmüller \footnotemark[5]
- name: Martin M. Gossner \footnotemark[8]
- name: Anton Güntsch \footnotemark[5]
- name: Malte Jochum \footnotemark[4]
- name: Birgitta König-Ries\footnotemark[6]
- name: Gaëtane Le Provost \footnotemark[2]
- name: Pete Manning \footnotemark[2]
- name: Andreas Ostrowski \footnotemark[6]
- name: Caterina Penone \footnotemark[4]
- name: Nadja K. Simons \footnotemark[7]
csl: mee.csl
fontsize: 11pt
geometry: margin=1.4in
documentclass: scrartcl
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
bibliography: div05.bib
---

@@

\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com, c/o Birgitta König-Ries
Friedrich-Schiller-Universität Jena, Institut f. Informatik, Ernst-Abbe-Platz 2, 07743 Jena
}
\footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
\footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
\footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
\footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
\footnotetext[7]{Department of Ecology and Ecosystem Management, Technische Universität München, Freising, Germany}

@@

1. Trait-based approaches are widespread throughout ecological research, as they offer great potential to achieve a general understandnig of a wide range of ecological and evolutionary mechanisms. Accordingly, a wealth of trait data is available for many organism groups, but this data is underexploited due to a lack of standardisation and heterogeneity in data formats and definitions.
2. We review current initiatives and structures developed for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories. 
3. In order to facilitate the standardisation and harmonisation of distributed trait datasets by data providers and data users, we propose a standardised vocabulary that can be used for storing and sharing ecological trait data. We discuss potential incentives and challenges for the wide adoption of such a standard by data providers. 
4. The use of a standard vocabulary allows trait datasets from heterogeneous sources to be compiled more easily into databases and facilitates the creation of interfaces between software tools for trait-data handling and analysis. By aiding decentralised trait-data standardisation, our data standard may ease data integration and use of trait data for a broader ecological research community and enable global syntheses across a wide range of taxa and ecosystems.

@@

functional ecology, species traits, semantic web, ontologies, data standard, ecoinformatics

@@

Functional traits are phenotypic (i.e. morphological, physiological, behavioral) characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. 
Recent years have seen a proliferation of trait-based research in a wide range of fields: trait data have been used to understand the evolutionary basis of individual-level properties [@salguero-gomez16], global patterns of biodiversity [@diaz16] and the relationship between ecosystem functions and the functional composition of species assemblages [@mouillot13; @perovic15; @bello10]. This research provides the mechanistic framework for linking climate change or anthropogenic land use to biodiversity and its related functions [@diaz11; @lavorel12; @allan15].
Species traits have been suggested as indicator variables for monitoring  ecosystem health at the individual level, like for instance changes in average body sizes in a population of fish [@kissling18]. 
 <!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
Using traits is also a promising means of bypassing taxonomic impediment, i.e. the fact that most species are yet undescribed and little is known of their interactions with other organisms and their environment. This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@leprovost17; @duarte11; @schrodt15].

@@

<!-- 2. trait databases have been developed, with standard structure and terms --> 
Despite the importance of functional trait data, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa as well as in evolutionary deep-time. 
However, the heterogeneity of data arising from different research contexts render trait data extremely heterogeneous and make the task of data compilation time-consuming and error-prone.
<!-- i.e. differences in taxonomic nomenclature and resolution (e.g. reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements--> 
To date, trait data have traditionally been harmonised and compiled into centralised databases only for specific organism groups and regional scope, often centered around particular research questions [e.g. PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17]. Less well-studied taxa and specialised research questions lack the resources for such an endeavour.
Besides initiatives aiming at assembling data, tools to enable the compatibility of data across databases are being developed. These include software to access trait data from the internet [e.g. @chamberlain17; @ankenbrand18], semantic-web standards [@wieczorek12; @page08] and thesauri of consensus terms [@walls12; @garnier17].

@@

Identifier (ID) | A unique label that relates row entries within and across *datasets*; may be used to connect multiple *data tables* into a *database*; can be user-specific or, in form of a *URI*, point to a globally valid *ontology* or *thesaurus*  .
Metadata | Data documentation of the higher level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97].
Occurrence | The observation context of a single entity of a taxon, i.e. data on an individual observed at a particular place and time. (http://rs.tdwg.org/dwc/terms/Organism); Sometimes used as synonym of 'observation' in data management context.

@@

Thesaurus | *Controlled vocabulary* that provides key *terms* with their associated *concepts* and relations for a specific field or domain of interest [@laporte13]; e.g. may define a hierarchy of broader or narrower terms.

@@

We consider these initiatives separately although they are often developed in conjunction to serve a particular database project, such as the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Traits [TOP; @garnier17]. 
We show how the degree of trait-data standardisation in existing datasets is highly variable, and which tools and standards are currently applied to achieve harmonisation of data from multiple, distributed sources. 
The objective of this review is to raise awareness of the generic structure of trait data and aid researchers in how to share and publish their own datasets in an appropriate form.

@@

In the field of comparative biology, morphological traits, such as traits  related to flower shape, leaf and stem structures for plants or wing and beak measurements for birds, as well as life-history traits such as Ellenberg values for plants or physiological and reproductive traits for animals (e.g. feeding biology, dispersal, metabolic rates and body size) have been assessed for decades and have been published in regular journal articles or books. 
With the rise of ecological trait-based research, measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits. 
A plethora of such static datasets has been published alongside scientific articles, or as standalone data publications [see @kleyer08 for a review on plant data; for animal data, see e.g., @gossner15; see Supplementary Material A, Table A1].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->

Today, the online publication of such data is greatly facilitated by file hosting services (e.g. Figshare, Zenodo, Researchgate, Data Dryad), which warrant long-term accessibility, and citeability via DOIs, and govern data sharing via license statements.
These platforms offer the hosting of publicly accessible *file repositories* at low-cost or for free, which makes them attractive for small and intermediate-sized research projects that cannot dedicate extra resources for data management. 
However, although open for download, the trait datasets on data repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make extraction tedious and dependent on specific commercial software, putting the long-term and open accessibility of these data at risk. 
Most importantly, these platforms enable public hosting of data with very low quality thresholds regarding *metadata* documentation and data standardisation. As a result, published datasets greatly vary in structure, nomenclature and metadata quality.

For trait data, there are common issues arising from the variability of data structures. 
For instance, the column descriptions and terminology applied to taxa and traits are mostly project specific and rarely chosen to allow translation into larger database initiatives. Furthermore, the metadata vary in their level of detail, e.g. for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a]. 
In terms of structure, trait data usually are reported in a species$\times$traits wide-table format. In this intuitive data table, each row represents a species (or taxon) for which multiple traits are reported in columns. <!-- Also, sources of  measurement bias are documented (methodology, observer ID, n of observations).  -->
Similarly, when reporting raw data, researchers place observations of individual organisms (i.e. the *occurrences*) in rows with multiple trait measurements applied to the same individual across multiple columns. 
Co-variates on the taxon, the individual specimen (e.g. sex or lifestage) or context of observation (e.g. time and place of sampling) would be placed in additional columns, and would further expand the two-dimensional data table. The resolution or scope of these co-variates varies greatly depending on the research question and observation context. 
Variability in the number and meaning of columns in these data tables requires tedious manual adjustments when merging multiple datasets [@wickham14]. A task that is taken up by data curators, who compile heterogeneous data into comprehensive databases (see next section).

@@

<!-- community database backend, but static publication --> 
In the past two decades, many distributed trait datasets have been aggregated and harmonised into greater collections with particular taxonomic or regional focus [e.g. @kleyer08; @oliveira17, see Supplementary Material A, Table A1].
While mostly concerned with issues of heterogeneity in units or levels of categorical variables, and aiming for high taxonomic coverage, few of these datasets apply a standardised terminology for taxa or traits that would allow them to be efficiently linked to other datasets.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary. -->
Documentation of metadata and methodology differs in the level of detail, depending on the research focus. 
Trait data compilations are often labelled 'database', although they do not formally provide data in a database structure in the strict sense of biodiveristy informatics.  
Instead, the data are released as static data tables of raw measurements or aggregate trait values on journal websites or open-access file hosting platforms, which may be updated irregularly.

As they deal with much larger amounts of data, initiatives that compile data from natural history museum collections are traditionally more concerned with standardisation. 
The amount of morphological measurements data extracted from museum collections and herbaria is likely to skyrocket in the near future due to digitisation efforts [@vollmar10; @blagoderov12] supported by new technology for scanning and pattern recognition [@younis18; @mantle12; @strobel18] and citizen science initiatives (e.g. www.markmybird.org). 
For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data from collections [@guralnick16]. The resulting data are published as versioned data tables which are updated as new data sources become available (vertnet.org).

<!-- database backend, selective download, user access management --> 
Specialised online portals have been created to attract data submissions from a defined research field and take care of data harmonisation, thereby greatly facilitating data synthesis.
<!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
For example, by aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other trait-data platform. The online portal enables selective data download and user permission and rights management. As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section). 
For animal trait data, however, a single unified platform and harmonising scheme is still lacking. The fact that animals span multiple trophic levels and possess diverse body structures makes this a more complex task than for plants [@moretti17].

@@

A *thesaurus* provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17]. 
In thesauri, hierarchies of concepts can be formalized by linking each term to broader or narrower terms, or to synonyms. For example, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'. 
Being publicly available, it is also possible to refer to these defined terms via globally unique *Uniform Resource Identifiers (URIs)*. For example, a measurement of fruit mass could be linked to the definition of the term within the TOP Thesaurus of plant traits [@garnier17] via its URI ‘http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass’.
*Ontologies* link defined terms by formally describing the relationships between them and providing a semantic model of the concepts used within a domain of research, with the objective of enabling the computational interpretation of data [@williams06; @walls12; @walls14; @kissling18]. 
Ontologies define terms based on other well-defined terms from published ontologies. The Plant Trait Ontology (TO) definition of the concept 'seed size' contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)." 
Furthermore, trait definitions may refer to related terms or synonyms defined in other trait ontologies or other scientific ontologies, like units as defined by the Units of Measurement Ontology [@gkoutos12]. 
By providing ontologies in a formalized file format, like OWL or RDF, a machine-readable web of definitions is spun across the Internet allowing researchers and search engines to relate independent trait measurements with each other and connect them to the wider *semantic web* of online data [@berners-lee01; @page08; @williams06; @walls12].

Comprehensive trait thesauri have been developed in the TOP Thesaurus of Plant Traits, which is employed in the TRY database [@garnier17], and in the Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14]. 
Ontologies of trait definitions have been developed for plants [e.g. the Plant Ontology, @jaiswal05; the Flora Phenotype Ontology, @hoehndorf16], as well as for animals [e.g. the Hymenoptera Anatomy Ontology, @yoder10; the Vertebrate Trait Ontology, @park13]. 
The existing thesauri and ontologies for traits differ widely in terms of hierarchical depth and detail, as well as in curation efforts and measures for peer-reviewed quality control. 
<!--Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g. units or legit factor levels) of measured values or reported facts. It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits. --><!-- BKR: Isn’t that the same? Or can you have multiple parents for one node? FDS: agree. simplify phrasing. -->

To conclude, there is already a suite of globally available thesauri and ontologies for traits available. However, definitions in some domains are better covered than others [@kissling18] and the interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards. To this end, the OBO Foundry is providing a development platform for (biological) ontologies and offers review and quality control [@smith07, http://www.obofoundry.org/]. 
While defined vocabularies are increasingly used in biodiversity data management, distributed trait-data of smaller projects published in general-purpose file servers rarely refer to standard terminologies. 
Finding and applying the most suited and highest quality ontology from the range of available ontologies is not an easy task for ecological researchers. 
To mitigate this effort, meta-ontology initiatives, like Ontobee (http://www.ontobee.org/), Bioportal (https://bioportal.bioontology.org/, @whetzel11), or the GFBio Terminology Service [@karam16, https://terminologies.gfbio.org/], provide centralised hosting for trait ontologies, structured browsing, and harmonized web services for computational access.

@@

To make sense of trait data in the context of more general databases, a consensus definition of trait data has been provided by  @garnier17: 
The define trait data to follow an entity-quality model (EQ), where a trait is defined as 'an entity having a quality'. 
More specifically, a trait dataset contains information on quantitative measurement* or qualitative facts (i.e. trait values) describing the physical phenotypic characteristics relating to fitness and performance (i.e. traits) observed on a biological entity (i.e. an individual specimen, or its parts) that can be assigned to a biological taxon (i.e. a species or higher-level taxon). 
The entity at which the trait was assessed may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g. a genus).

These relationships between a trait measurement and an individual organism as an entity of a particular taxon have been formalised in the schema for biological collection records [Access to Biological Collection Data, ABCD; @holetschek12] or the Darwin Core Standard for biodiversity data [DwC; @wieczorek12]. 
These frameworks specify terms and classes to describe the general structure of biodiversity databases. The terms may be used for  defining columns in data tables that contain measurement values, units and categorical levels, taxon names, variables such as sex or life stage, ancillary information of time and date of observation, and methodological details. The terminologies provided by these standards are quite universal and would even cover most use cases of trait data. 
A suite of terminology extensions links to and expands the capacities of DwC [@wieczorek12].
For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) uses DwC terms. 
Both DwC and ABCD are ratified standards of the Biodiversity Information Standards TDWG (Taxonomic Databases Working Group; www.tdwg.org) which is a global network to support the development and wide adoption of exchange standards for biodiversity data.

@@

## Closing gaps in the trait-data lifecycle

In sum, we attest a gap between the trait-data structures developed for data curators and managers and the data input produced by data providers. Hardly any of the aforementioned standalone or aggregated trait datasets for birds, amphibians, mammals or invertebrates employs the described standard terminologies, ontologies or semantic web standards. As it stands, re-using these data in larger compilations or integrating them into structured database initiatives is error-prone and labour-intensive and the potential for a broad synthesis is diminished.

>>>>

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1]  \footnotemark[2]
- name: David Fichtmüller \footnotemark[5]
- name: Martin M. Gossner \footnotemark[8]
- name: Anton Güntsch \footnotemark[5]
- name: Malte Jochum \footnotemark[4]
- name: Birgitta König-Ries\footnotemark[6]
- name: Gaëtane Le Provost \footnotemark[2]
- name: Pete Manning \footnotemark[2]
- name: Andreas Ostrowski \footnotemark[6]
- name: Caterina Penone \footnotemark[4]
- name: Nadja K. Simons \footnotemark[7] \footnotemark[9]
csl: mee.csl
fontsize: 11pt
geometry: margin=1.4in
documentclass: scrartcl
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  word_document:
    fig_caption: yes
    fig_width: 6
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
bibliography: div05.bib
---

@@

\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com, c/o Birgitta König-Ries
Friedrich-Schiller-Universität Jena, Institut f. Informatik, Ernst-Abbe-Platz 2, 07743 Jena
}
\footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
\footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
\footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
\footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
\footnotetext[7]{Department of Ecology and Ecosystem Management, Technische Universität München, Freising, Germany}
\footnotetext[9]{Ecological Networks, Technische Universität Darmstadt, Schnittspahnstraße 3, 64287 Darmstadt, Germany}

@@

1. Trait-based approaches are widespread throughout ecological research as they offer great potential to achieve a general understanding of a wide range of ecological and evolutionary mechanisms. Accordingly, a wealth of trait data is available for many organism groups, but this data is underexploited due to a lack of standardisation and heterogeneity in data formats and definitions.
2. We review current initiatives and structures developed for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories. 
3. In order to facilitate the standardisation and harmonisation of distributed trait datasets by data providers and data users, we propose a standardised vocabulary that can be used for storing and sharing ecological trait data. We discuss potential incentives and challenges for the wide adoption of such a standard by data providers. 
4. The use of a standard vocabulary allows trait datasets from heterogeneous sources to be aggregated more easily into compilations and facilitates the creation of interfaces between software tools for trait-data handling and analysis. By aiding decentralised trait-data standardisation, our vocabulary may ease data integration and use of trait data for a broader ecological research community and enable global syntheses across a wide range of taxa and ecosystems.

@@

functional ecology, species traits, semantic web, ontologies, data standardisation, ecoinformatics

@@

Functional traits are phenotypic (i.e. morphological, physiological, behavioral) characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. 
Recent years have seen a proliferation of trait-based research in a wide range of fields: trait data have been used to understand the evolutionary basis of individual-level properties [@salguero-gomez16], global patterns of biodiversity [@diaz16] and the relationship between ecosystem functions and the functional composition of species assemblages [@mouillot13; @perovic15; @bello10]. This research provides the mechanistic framework for linking climate change or anthropogenic land use to biodiversity and its related functions [@diaz11; @lavorel12; @allan15].
Species traits have been suggested as indicator variables for monitoring  ecosystem health at the individual level, like for instance changes in average body sizes in a population of fish [@kissling18]. 
<!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
Using traits is also a promising means of bypassing taxonomic impediment, i.e. the fact that most species are yet undescribed, and little is known of their interactions with other organisms and their environment. This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@grime01; @moretti17; @villeger17].

@@

<!-- 2. trait databases have been developed, with standard structure and terms --> 
Despite the importance of functional trait data, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa as well as in evolutionary deep-time. 
However, the heterogeneity of data arising from different research contexts render trait data extremely heterogeneous and make the task of data compilation time-consuming and error-prone.
<!-- i.e. differences in taxonomic nomenclature and resolution (e.g. reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements-->
To date, trait data have traditionally been harmonised and compiled into centralised databases only for specific organism groups and regional scope, often centered around particular research questions [e.g., PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17]. Less well-studied taxa and specialised research questions lack the resources for such an endeavour.
Besides initiatives aiming at assembling data, tools to enable the compatibility of data across databases are being developed. These include software to access trait data from the internet [e.g., @chamberlain17; @ankenbrand18], semantic-web standards [@wieczorek12; @page08] and thesauri of consensus terms [@walls12; @garnier17].

@@

Identifier (ID) | A unique label that relates data entries to information within and across *datasets* or external items of information; may be used to connect multiple *data tables* into a *database*; can be user-specific or, in form of a *URI*, point to a globally valid *ontology* or *thesaurus*.
Metadata | Data documentation of the higher-level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97].
Occurrence | The observation context of a single individual, i.e. the existance of an organism at a particular place and time; Sometimes used as synonym of 'observation' in data management context.

@@

Thesaurus | *Controlled vocabulary* that provides key *terms* with their associated *concepts* and relations for a specific field or domain of interest [@laporte13]; e.g., may define a hierarchy of broader or narrower terms.

@@

We consider these initiatives separately although they are often developed in conjunction to serve a particular database project, such as the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Characteristics [TOP; @garnier17]. 
We show how the degree of trait-data standardisation in existing datasets is highly variable, and which tools and standards are currently applied to achieve harmonisation of data from multiple, distributed sources. 
The objective of this review is to raise awareness of the generic structure of trait data and aid researchers in how to share and publish their own datasets in an appropriate form.

@@

In the field of comparative biology, morphological traits, such as traits  related to flower shape, leaf and stem structures for plants or wing and beak measurements for birds, as well as life-history traits such as Ellenberg values for plants or physiological and reproductive traits for animals (e.g., feeding biology, dispersal, metabolic rates and body size) have been assessed for decades and have been published in regular journal articles or books. 
With the rise of ecological trait-based research, measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits. 
A plethora of such static datasets has been published alongside scientific articles, or as standalone data publications [see @kleyer08 for a review on plant data; for animal data, e.g., @gossner15 and Supplementary Material A, Table A1].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->

Today, the online publication of such data is greatly facilitated by file hosting services (e.g., Figshare, Zenodo, Researchgate, Data Dryad), which warrant long-term accessibility, and citeability via DOIs, and govern data sharing via license statements.
These platforms offer the hosting of publicly accessible *file repositories* at low-cost or for free, which makes them attractive for small and intermediate-sized research projects that cannot dedicate extra resources for data management. 
<!--However, although open for download, the trait datasets on data repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make extraction tedious and dependent on specific commercial software, putting the long-term and open accessibility of these data at risk. -->
Most importantly, these platforms enable public hosting of data with very low quality thresholds regarding *metadata* documentation and data standardisation. Thus, although open for download, the trait datasets on data repositories might be stored in variable tabular structures and labelled following self-defined terms which makes extraction and further use unnecessarily tedious. <!--As a result, published datasets greatly vary in structure, nomenclature and metadata quality. -->

For trait data, there are common issues arising from the variability of data structures. 
For instance, the column descriptions and terminology applied to taxa and traits are mostly project specific and rarely chosen to allow translation into larger database initiatives. Furthermore, the metadata vary in their level of detail, e.g., for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a]. 
In terms of structure, trait data usually are reported in a species$\times$traits wide-table format. In this intuitive data table, each row represents a species (or taxon) for which multiple traits are reported in columns.<!-- Also, sources of  measurement bias are documented (methodology, observer ID, n of observations).  -->
Similarly, when reporting raw data, researchers place observations on individual organisms in rows with multiple trait measurements applied to the same individual across multiple columns. 
Co-variates on the taxon, the individual specimen (e.g., sex or lifestage) or context of observation (e.g., time and place of sampling) would be placed in additional columns and would further expand the two-dimensional data table. The resolution or scope of these co-variates varies greatly depending on the research question and observation context. 
Variability in the number and meaning of columns in these data tables requires tedious manual adjustments when merging multiple datasets [@wickham14]. A task that is taken up by data curators, who compile heterogeneous data into comprehensive databases (see next section).

@@

<!-- community database backend, but static publication --> 
In the past two decades, many distributed trait datasets have been aggregated and harmonised into greater collections with particular taxonomic or regional focus [e.g., @kleyer08; @oliveira17, see Supplementary Material A, Table A1].
While these initiatives address issues of heterogeneity in units or categorical variables, or aim for high taxonomic or geographic coverage, few of these compilations apply a standardised terminology for taxa or trait definitions.
Additionally, documentation of metadata and methodology in the original datasets often differs in the level of detail, depending on the research focus, and those information are difficult to maintain in the compilation.
Particularly, in the process of data aggregation, rich metadata content might be lost or the reference to the original dataset might be obscured, as only aggregated values are reported (e.g. means or medians). 
Such trait data compilations are often labelled 'database', although they do not formally provide data in a database structure in the strict sense of biodiversity informatics.
Instead, the data are released as static data tables of raw measurements or aggregate trait values on journal websites or open-access file hosting platforms, which may be updated irregularly.

As they deal with much larger amounts of data, initiatives that compile data from natural history museum collections are traditionally more concerned with standardisation. 
The amount of morphological measurements data extracted from museum collections and herbaria is likely to skyrocket in the near future due to digitisation efforts [@vollmar10; @blagoderov12] supported by new technology for scanning and pattern recognition [@younis18; @mantle12; @strobel18] and citizen science initiatives (e.g., www.markmybird.org). 
For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data from collections [@guralnick16]. The resulting data are published as versioned data tables which are updated as new data sources become available (vertnet.org).

<!-- database backend, selective download, user access management --> 
Specialised online portals have been created to attract data submissions from a defined research field and take care of data harmonisation, thereby greatly facilitating data synthesis.
For example, by aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other trait-data platform. The online portal enables selective data download and user permission and right management.<!--As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section). -->
For animal trait data, however, a single unified platform and harmonising scheme is still lacking. <!--The fact that animals span multiple trophic levels and possess diverse body structures makes this a more complex task than for plants [@moretti17].-->

@@

A *thesaurus* provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17]. 
To provide a logic structure for trait terms, Garnier et al. (2017) suggest the Entity-Quality model (EQ), where a trait is defined as ‘an entity having a quality’ (for instance for trait ‘femur length’, ‘femur’ is the entity and ‘length’ the quality, Fig. 1).
In thesauri, hierarchies of concepts can be formalized by linking each term to broader or narrower terms, or to synonyms. For example, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'. 
Being publicly available, it is also possible to refer to these defined terms via globally unique *Uniform Resource Identifiers (URIs)*. For example, a measurement of fruit mass could be linked to the definition of the term within the Thesaurus of Plant Characteristics [TOP, @garnier17] via its URI ‘http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass’.
In addition to defining terms for human interpretation, *ontologies* define terms by their relationship to other defined terms, thereby providing a semantic model of the concepts used within a domain of research, with the objective of enabling the computational interpretation of data [@williams06; @walls12; @walls14; @kissling18]. 
 The Plant Trait Ontology (TO) definition of the concept 'seed size' contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)." 
Furthermore, trait definitions may refer to related terms or synonyms defined in other trait ontologies or other scientific ontologies, like units as defined by the Units of Measurement Ontology [@gkoutos12]. 
By providing ontologies in a formalized syntax, like Web Ontology Language (OWL), a machine-readable web of definitions is spun across the Internet allowing researchers and search engines to relate independent trait measurements with each other and connect them to the wider *semantic web* of online data [@berners-lee01; @page08; @williams06; @walls12].

Comprehensive trait thesauri have been developed in TOP [which is employed in the TRY database, @garnier17] and in the Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14]. 
Ontologies of trait definitions have been developed for plants [e.g., the Plant Ontology, @jaiswal05; the Flora Phenotype Ontology, @hoehndorf16], as well as for animals [e.g., the Hymenoptera Anatomy Ontology, @yoder10; the Vertebrate Trait Ontology, @park13]. 
The existing thesauri and ontologies for traits differ widely in terms of hierarchical depth and detail, as well as in curation efforts and measures for peer-reviewed quality control. 
<!--Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g. units or legit factor levels) of measured values or reported facts. It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits. --><!-- BKR: Isn’t that the same? Or can you have multiple parents for one node? FDS: agree. simplify phrasing. -->

To conclude, there is already a suite of globally available thesauri and ontologies for traits. However, definitions in some domains are better covered than others [@kissling18] and the interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards. To this end, the OBO Foundry is providing a development platform for (biological) ontologies and offers review and quality control [@smith07, http://www.obofoundry.org/]. 
While defined vocabularies are increasingly used in biodiversity data management, distributed trait-data of smaller projects published in general-purpose file servers rarely refer to standard terminologies. 
Finding and applying the most suited and highest quality ontology from the range of available ontologies is not an easy task for ecological researchers. 
To mitigate this effort, meta-ontology initiatives, like Ontobee (http://www.ontobee.org/), Bioportal (https://bioportal.bioontology.org/, @whetzel11), or the GFBio Terminology Service [@karam16, https://terminologies.gfbio.org/], provide centralised hosting for trait ontologies, structured browsing, and harmonized web services for computational access.

@@

<!--To make sense of trait data in the context of more general databases, a consensus definition of trait data has been provided by  @garnier17: 
The define trait data to follow an entity-quality model (EQ), where a trait is defined as 'an entity having a quality'. 
More specifically, a trait dataset contains information on quantitative measurement* or qualitative facts (i.e. trait values) describing the physical phenotypic characteristics relating to fitness and performance (i.e. traits) observed on a biological entity (i.e. an individual specimen, or its parts) that can be assigned to a biological taxon (i.e. a species or higher-level taxon). 
The entity at which the trait was assessed may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g. a genus).-->
In terms of structure, a trait dataset contains measurement values or qualitative facts reporting on traits observed from one or multiple organisms or occurrences, i.e. observa-tions of an individual at a specific place and time, that are assigned to a scientific taxon. Information is thus centered around an observed numerical or categorical value, being further specified in what has been measured (i.e., a trait) and on what instance of a sci-entific taxon it has been observed. Additional information would describe the situation of the measurement, like methods of data aggregation, technical tools or observer identi-ty, and context of the occurrence, like environmental factors at the time and place of sampling or growth conditions. Depending on the research question, these co-variates are used to explain noise or drivers in trait values.

Standard terms for the formal description of the common concepts of biodiversity knowledge have been provided in the schema for biological collection records [Access to Biological Collection Data, ABCD; @holetschek12] or the Darwin Core Standard for biodiversity data [DwC; @wieczorek12]. 
These frameworks specify terms and classes to describe the describe the relationships between any kind of measurement and occurrence of organisms of a particular taxon. The terms may be used for  defining columns in data tables that contain measurement values, units and categorical levels, taxon names, variables such as sex or life stage, ancillary information of time and date of observation, and methodological details. The terminologies provided by these standards are quite universal and cover most use cases of trait data. 
A suite of terminology extensions links to and expands the capacities of DwC [@wieczorek12].
<!--For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) uses DwC terms. -->
Both DwC and ABCD are ratified standards of the Biodiversity Information Standards TDWG (Taxonomic Databases Working Group; www.tdwg.org) which is a global network to support the development and wide adoption of exchange standards for biodiversity data.

@@

## Closing gaps to improve trait-data re-use

In sum, we attest to a gap between the trait-data structures developed for data curators and managers and the data input produced by data providers. Hardly any of the aforementioned standalone or aggregated trait datasets for birds, amphibians, mammals or invertebrates employs the described standard terminologies, ontologies or semantic web standards. As it stands, re-using these data in larger compilations or integrating them into structured database initiatives is error-prone and labour-intensive and the potential for a broad synthesis is diminished.
